---
title: "Practical Machine Learning: Prediction Assignment"
author: "Yasser Gonzalez -- http://yassergonzalez.com"
date: "September 2015"
output: html_document
---

# Introduction

The goal of this assignment is to build a prediction model capable of
identifying if a subject is performing a specific weight lifting
exercise correctly.

A data set is available with information collected from six subjects
who were asked to perform 10 repetitions of dumbbell lifts in five
different settings:

1.   exactly according to the specification (class `A`),
2.   throwing the elbows to the front (class `B`),
3.   lifting the dumbbell only halfway (class `C`),
4.   lowering the dumbbell only halfway (class `D`), and
5.   throwing the hips to the front (class `E`).

The model's input consists of a number of sensors attached to the body
of the subject: accelerometers on the belt, forearm, arm, and dumbell.
Using this information, the model must provide a decision about the
correctness of the exercise execution according to one of the five
categories specified above.

Additional information about the experimental setup and the data set
is available at http://groupware.les.inf.puc-rio.br/har and in the
following publication:

> Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H.
> [Qualitative Activity Recognition of Weight Lifting Exercises. ](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201)
> *4th International Conference in Cooperation with SIGCHI (Augmented Human '13)*.
> Stuttgart, Germany: ACM SIGCHI, 2013.

# Initialization

The following code fragment loads the required R packages and sets the
random seed to ease the reproducibility of the results.

```{r results='hide', message=FALSE, warning=FALSE}
library("readr")
library("dplyr")
library("ggplot2")
library("caret")
library("randomForest")
library("doMC")

set.seed(2100479551)
```

# Data Loading and Feature Selection

```{r warning = FALSE}
dataset_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
dataset_csv <- "pml-training.csv"
if (!file.exists(dataset_csv)) {
    download.file(dataset_url, dataset_csv, method = "curl")
}
dataset <- read_csv(dataset_csv)
```

```{r}
sensors <- c("forearm", "arm", "belt", "dumbbell")
vars <- c("roll_%s", "pitch_%s", "yaw_%s",
          "gyros_%s_x", "gyros_%s_y", "gyros_%s_z",
          "accel_%s_x", "accel_%s_y", "accel_%s_z",
          "magnet_%s_x", "magnet_%s_y", "magnet_%s_z")
features <- unlist(lapply(sensors,
      function (sensor) lapply(vars,
               function (var) sprintf(var, sensor))))
```

```{r}
dataset <- dataset %>%
    select(one_of(c("classe", features))) %>%
    mutate(classe = as.factor(classe))
```

```{r}
indices <- createDataPartition(dataset$classe, p = 0.8, list = FALSE)
training_data <- dataset[indices, ]
testing_data <- dataset[-indices, ]
```

# Exploratory Data Analysis

```{r}
```

# Building the Prediction Model

```{r}
model_file <- "model.rds"
if (!file.exists(model_file)) {
    registerDoMC(cores = 4)
    model <- train(classe ~ ., data = training_data,
                   method = "rf",
                   tuneLength = 8, # number of mtry values
                   metric = "Accuracy",
                   trControl = trainControl(method = "cv", number = 10))
    saveRDS(model, model_file)
} else {
    model <- readRDS(model_file)
}
```

# Evaluating the Prediction Model

```{r}
confusionMatrix(predict(model, newdata = testing_data),
                testing_data$classe)
```

# Submission Results

```{r}
testing_dataset_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testing_dataset_csv <- "pml-testing.csv"
if (!file.exists(testing_dataset_csv)) {
    download.file(testing_dataset_url, testing_dataset_csv, method = "curl")
}

testing_dataset <- read_csv(testing_dataset_csv) %>%
    select(one_of(c("problem_id", features)))
```

```{r}
prediction <- predict(model, newdata = testing_dataset)
prediction
```

```{r}
for (i in seq(along = prediction)) {
    txt_file <- paste0("problem_id_", testing_dataset$problem_id[i], ".txt")
    write.table(prediction[i], file = txt_file,
                quote = FALSE, row.names = FALSE, col.names = FALSE)
}
```
