---
title: "Practical Machine Learning: Prediction Assignment"
author: "Yasser Gonzalez -- http://yassergonzalez.com"
date: "September 2015"
output: html_document
---

# Introduction

The goal of this assignment is to build a prediction model capable of
identifying if a subject is performing a specific weight lifting
exercise correctly.

A data set is available with information collected from six subjects
who were asked to perform 10 repetitions of dumbbell lifts in five
different settings:

1.   exactly according to the specification (class `A`),
2.   throwing the elbows to the front (class `B`),
3.   lifting the dumbbell only halfway (class `C`),
4.   lowering the dumbbell only halfway (class `D`), and
5.   throwing the hips to the front (class `E`).

The model's input consists of a number of sensors attached to the body
of the subject: on the belt, forearm, arm, and dumbell. Using this
information, the model must provide a decision about the correctness
of the exercise execution according to one of the five categories
specified above.

Additional information about the experimental setup and the data set
is available at http://groupware.les.inf.puc-rio.br/har and in the
following publication:

> Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H.
> [Qualitative Activity Recognition of Weight Lifting Exercises. ](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201)
> *4th International Conference in Cooperation with SIGCHI (Augmented Human '13)*.
> Stuttgart, Germany: ACM SIGCHI, 2013.

# Initialization

The following code fragment loads the required R packages and sets the
random seed to ease the reproducibility of the results.

```{r results='hide', message=FALSE, warning=FALSE}
library("readr")
library("dplyr")
library("ggplot2")
library("caret")
library("randomForest")
library("doMC")

set.seed(2100479551)
```

# Data Loading and Feature Selection

First, the data set is downloaded and loaded into R.

```{r warning = FALSE}
dataset_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
dataset_csv <- "pml-training.csv"
if (!file.exists(dataset_csv)) {
    download.file(dataset_url, dataset_csv, method = "curl")
}
dataset <- read_csv(dataset_csv)
```

In addition to the different readings from the sensors, the data set
includes a number of engineered features (such as mean and standard
deviation of the sensor readings computed over a certain sliding time
window), but for this assignment I have decided to work with the
original sensor readings and expand the set of features later if
deemed necessary. Therefore, the selected features correspond to the
Euler angles (roll, pitch and yaw), and the 3-axis coordinates of the
accelerometer, gyroscope and magnetometer readings for each one of the
four sensors.

```{r}
sensors <- c("forearm", "arm", "belt", "dumbbell")
vars <- c("roll_%s", "pitch_%s", "yaw_%s",
          "gyros_%s_x", "gyros_%s_y", "gyros_%s_z",
          "accel_%s_x", "accel_%s_y", "accel_%s_z",
          "magnet_%s_x", "magnet_%s_y", "magnet_%s_z")
features <- unlist(lapply(sensors,
      function (sensor) lapply(vars,
               function (var) sprintf(var, sensor))))

dataset <- dataset %>%
    select(one_of(c("classe", features))) %>%
    mutate(classe = as.factor(classe))
```

There are `r nrow(dataset)` observations of the `r length(features)`
features in the data set.

Before continuing, the data set is divided in two parts: 80% of the
available samples will be used to train the prediction model and the
remaining 20% for testing. The testing portion of the data set will be
used to estimate the out-of-sample error of the prediction model.

```{r}
indices <- createDataPartition(dataset$classe, p = 0.8, list = FALSE)
training_data <- dataset[indices, ]
testing_data <- dataset[-indices, ]
```

The training data set has `r nrow(training_data)` samples and the
training data set `r nrow(testing_data)`.

# Building the Prediction Model

```{r}
```

```{r}
model_file <- "model.rds"
if (!file.exists(model_file)) {
    registerDoMC(cores = 4)
    model <- train(classe ~ ., data = training_data,
                   method = "rf",
                   tuneLength = 8, # number of mtry values
                   metric = "Accuracy",
                   trControl = trainControl(method = "cv", number = 10))
    saveRDS(model, model_file)
} else {
    model <- readRDS(model_file)
}
```

# Evaluating the Prediction Model

```{r}
confusionMatrix(predict(model, newdata = testing_data),
                testing_data$classe)
```

# Submission Results

```{r}
testing_dataset_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testing_dataset_csv <- "pml-testing.csv"
if (!file.exists(testing_dataset_csv)) {
    download.file(testing_dataset_url, testing_dataset_csv, method = "curl")
}

testing_dataset <- read_csv(testing_dataset_csv) %>%
    select(one_of(c("problem_id", features)))
```

```{r}
prediction <- predict(model, newdata = testing_dataset)
prediction
```

```{r}
for (i in seq(along = prediction)) {
    txt_file <- paste0("problem_id_", testing_dataset$problem_id[i], ".txt")
    write.table(prediction[i], file = txt_file,
                quote = FALSE, row.names = FALSE, col.names = FALSE)
}
```
